Computing with an Angry Poker Player

### 2014-09-25 draft0.tex
adding bounds until stable section

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\setlength{\parindent}{0cm}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\setlength{\parindent}{0cm}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\section{Time until Stability}

We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state. We will first focus on the case of the board with only one stack containing $n$ chips. Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$. We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes. We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize. Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.

One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function. Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$. So we know that when the board is stable at some $f'$, 
\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
\[ T(n) = O(n^2) \]
We can get tighter bounds by looking at the number of earthquakes more carefully. Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse). In particular,
\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
 \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right. \]
So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
\begin{figure}[!ht]
\centering
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
\caption{Number of global earthquakes until stability}
\label{fig:growthofT}
\end{figure}
By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$. Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence, we believe that $T'(n) = \Theta(n\log n)$.

\begin{figure}[!ht]
\centering
\subfigure[Local earthquakes of center until stable. Horizontal axis is $n$ and vertical axis is the number of local earthquakes]{\includegraphics[width=0.4\linewidth]{centerearthquakes}\label{fig:case1}} \qquad
\subfigure[Local earthquakes of center until stable. Horizontal axis is $n$ and vertical axis is the number of local earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{centerearthquakesdividedbyn}\label{fig:case2}}
\caption{Number of local earthquakes in center until stability}
\label{fig:growthofT}
\end{figure}
